---
title: "Forecasting EURIBOR 12M with ARIMA using R"
output: github_document
---

```{r load_libraries, include=FALSE}
library(lubridate)
library(ggplot2)
library(dplyr)
library(tidyr)
library(forecast)
library(stringr)
library(here)

knitr::opts_chunk$set(echo = TRUE)
```

## Euribor Arima Forecast

Simple R Markdown that pulls historic 12M EURIBOR rates from https://www.euribor-rates.eu in JSON format and performs an ARIMA (Autoregressive integrated moving average) forecast for the next 180 days.

```{r load_historic_rates_from_csv, include=FALSE}

# Read historic EURIBOR rates already saved in CSV
csv_ds <- read.csv(here("euribor_rates.csv"), header = TRUE)

# Convert to data.frame
euribor <- data.frame(
  ymd(csv_ds$date),
  as.numeric(csv_ds$rate)
)
colnames(euribor) <- c("date", "rate")
```

```{r fetch_new_rates_from_external_api, include=FALSE}
# update with more recent rates
start <- max(euribor$date)
end <- Sys.Date()
while(start < end) {
  from <- start
  to <- from + years(1)
  
  if(to > end) 
    to = end
 
  mintick <- sprintf(as.numeric(as.POSIXct(from))*1000, fmt = '%.0f')
  maxtick <- sprintf(as.numeric(as.POSIXct(to))*1000, fmt = '%.0f')
  
  #print(paste0("Pulling:", from, " - ", to, "   (mintick,maxtick) = (",mintick,",",maxtick,")"))
  
  # generate the URL with the proper min and max ticks
  url <- paste0("https://www.euribor-rates.eu/umbraco/api/euriborpageapi/highchartsdata?minticks=",mintick,"&maxticks=",maxtick,"&series[0]=4")

  # fetch JSON
  historic_rates = jsonlite::fromJSON(txt=url)
  df <- data.frame(historic_rates$Data)
  
  # bind to historic rates
  new_rates <- data.frame(
    as.Date(as.POSIXct(df$X1/1000, origin="1970-01-01", tz = "UTC")),
    df$X2
  )
  colnames(new_rates) <- c("date", "rate")
  euribor <- rbind(
    euribor,
    new_rates
  )
  
  start <- to
  # sleep for X seconds to keep under the radar
  #Sys.sleep(15)
}
```

```{r update_historic_csv_with_new_rates, include=FALSE}
# remove duplicates
euribor <- unique(euribor)

# update historical rates csv
colnames(euribor) <- c("date", "rate")
write.csv(euribor, file=here("euribor_rates.csv"), row.names = FALSE)

```

## Preview Historical Rates Dataset
```{r historical_rates_head_and_tail, echo = FALSE}
# preview historical data.frame
head(euribor)
tail(euribor)
```

```{r train_arima_and_forecast, include=FALSE}
# train ARIMA
arima_fit = auto.arima(ts(euribor)[,2])
# Forecast for next X days
days_forecast <- 180
arima_forecast <- forecast(arima_fit, h=days_forecast)
# convert result for data.frame
euribor_forecasted <- data.frame(
  date <- seq(max(euribor$date)+1, by = "day", length.out = days_forecast),
  mean <- arima_forecast$mean,
  lo_80 <- arima_forecast$lower[,1],
  hi_80 <- arima_forecast$upper[,1],
  lo_95 <- arima_forecast$lower[,2],
  hi_95 <- arima_forecast$upper[,2]
)
colnames(euribor_forecasted) <- c("date", "forecast", "lo_80", "hi_80", "lo_95", "hi_95")

chartDates <- c(euribor$date, euribor_forecasted$date)
monthKeys <- my(paste0(str_pad(month(chartDates), 2, pad = "0"), '-', year(chartDates)))

chart_data <- data.frame(
  series <- c("Historic"),
  euribor$date,
  euribor$rate,
  NA,
  NA,
  NA,
  NA
)
colnames(chart_data) <- c("series", "date", "rate", "lo_80", "hi_80", "lo_95", "hi_95")

chart_data <- rbind(
  chart_data,
  data.frame(
    series = c("Forecast"),
    date = euribor_forecasted$date,
    rate = euribor_forecasted$forecast,
    lo_80 = euribor_forecasted$lo_80,
    hi_80 = euribor_forecasted$hi_80,
    lo_95 = euribor_forecasted$lo_95,
    hi_95 = euribor_forecasted$hi_95
  )
)

chart_data <- chart_data %>% arrange(desc(date))

min_chart_data <- min(c(chart_data$rate), min(chart_data$lo_80), min(chart_data$lo_95))
max_chart_data <- max(c(chart_data$rate), max(chart_data$hi_80), max(chart_data$hi_95))

```


## Preview Forecast Rates Dataset


```{r cars, echo = FALSE}
# preview data.frame
head(euribor_forecasted)
tail(euribor_forecasted)
```

```{r dates_for_text, include=FALSE, echo = FALSE}
last_historical_date <- max(euribor$date)
first_forecast_date <- min(euribor_forecasted$date)
last_forecast_date <- max(euribor_forecasted$date)

```

Last historic date is `r last_historical_date`, forecast range went from `r first_forecast_date` to `r last_forecast_date`.

## 1 Year Historical + Forecast with 80% & 95% confidence intervals
```{r one_historical_year_and_forecast_period, fig.width=11,fig.height=6, echo=FALSE}
ggplot(chart_data %>% filter(date >= max(euribor$date) - years(1)), aes(x=date, y = rate, color = series)) + 
  geom_line(aes(linetype=series), size = 1) +
  scale_linetype_manual(values=c("dashed", "solid")) +
  geom_ribbon(aes(x=date, ymin = lo_80, ymax = hi_80), color = "#4fc92e", size = 1, alpha = 0.1, linetype="dashed") +
  geom_ribbon(aes(x=date, ymin = lo_95, ymax = hi_95), color = "#c9a22e", size = 1, alpha = 0.1, linetype="dashed") +
  scale_y_continuous(limits = c(min_chart_data, max_chart_data), breaks = seq(from = -1, to = 6, by = 0.20), name ="Euribor") +
  scale_x_date(breaks = monthKeys[month(monthKeys) %in% c(1:12) & day(monthKeys) == 1], name ="Date") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
```

## All Historic Data + Forecast with 95% confidence interval
```{r all_historic_year_and_forecast_95_confidence_interval, fig.width=11,fig.height=6, echo=FALSE}
ggplot(chart_data, aes(x=date, y = rate, color = series)) + 
  geom_line(aes(linetype=series), size = 1) +
  scale_linetype_manual(values=c("dashed", "solid")) +
  geom_ribbon(aes(x=date, ymin = lo_95, ymax = hi_95), color = "#c9a22e", size = 1, alpha = 0.1, linetype="dashed") +
  scale_y_continuous(limits = c(min_chart_data, max_chart_data), breaks = seq(from = -1, to = 6, by = 0.20), name ="Euribor") +
  scale_x_date(breaks = monthKeys[month(monthKeys) %in% c(1) & day(monthKeys) == 1], name ="Date") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
```

## All Historic Data + Forecast with 80% confidence interval
```{r all_historic_year_and_forecast_80_confidence_interval, fig.width=11,fig.height=6, echo=FALSE}
ggplot(chart_data, aes(x=date, y = rate, color = series)) + 
  geom_line(aes(linetype=series), size = 1) +
  scale_linetype_manual(values=c("dashed", "solid")) +
  geom_ribbon(aes(x=date, ymin = lo_80, ymax = hi_80), color = "#4fc92e", size = 1, alpha = 0.1, linetype="dashed") +
  scale_y_continuous(limits = c(min_chart_data, max_chart_data), breaks = seq(from = -1, to = 6, by = 0.20), name ="Euribor") +
  scale_x_date(breaks = monthKeys[month(monthKeys) %in% c(1) & day(monthKeys) == 1], name ="Date") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
```

